### 2019-06-12 manuscript_clean.Rmd
Final versions for BioRxiv submission

---
bibliography: manuscript.bib
csl: systematic-biology.csl
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_document2:
    toc: FALSE
indent: TRUE
---

@@

Alejandro Damian-Serrano^1^^,‡^, Steven H.D. Haddock^2^, Elizabeth D. Hetherington^3^, C. Anela Choy^3^, Casey W. Dunn^1^

@@

^3^ Scripps Institution of Oceanography, University of California San Diego, 9500 Gilman Dr., La Jolla, CA 92093, USA

@@

Predators have evolved dedicated body parts to capture and subdue prey. As different predators specialize on distinct prey taxa, their tools for prey capture diverge into a variety of adaptive forms. Studying the evolution of predation is greatly facilitated by a predator clade with structures used exclusively for prey capture that present significant morphological variation. Siphonophores, a clade of colonial cnidarians, satisfy these criteria particularly well, capturing prey with their tentilla (tentacle side branches). Earlier work has shown that extant siphonophore diets correlate with the different morphologies and sizes of their tentilla and nematocysts. We hypothesize that evolutionary specialization on different prey types has driven the phenotypic evolution of these characters. To test this hypothesis, we: (1) measured multiple morphological traits from fixed siphonophore specimens using microscopy and high speed video techniques, (2) built a phylogenetic tree of 45 species, and (3) characterized the evolutionary associations between siphonophore nematocyst characters and prey type data from the literature. Our results show that siphonophore tentillum structure has strong evolutionary associations with prey type and size specialization, and suggest that shifts between prey-type specializations are linked to shifts in tentillum and nematocyst size and shape. In addition, we found that these morphological characters are useful for predicting siphonophore diets.  Thus, the evolutionary history of tentilla shows that siphonophores are an example of ecological niche diversification via morphological innovation and evolution. This study contributes to understanding how morphological evolution has shaped present-day oceanic food-webs.

@@

  Most animal predators have characteristic biological tools that they use to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures tend to be finely attuned to their ability to successfully capture specific prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (e.g. avoidance, escape, protective barriers) leads to modifications that can counter those defenses The more specialized the diet of a predator is, the more specialized its tools need to be to meet the specific challenges posed by the prey. Understanding the relationships between predatory specializations and morphological specializations is necessary to contextualize the phenotypic diversity of predators, and to quantify the importance of ecological diversification in generating this diversity.

  Siphonophores (Cnidaria : Hydrozoa) are a clade of organisms bearing modular structures that are exclusively used for prey capture: the tentilla (Fig. \@(figure1)). These present a significant morphological variation across species [@mapstone2014global] (Fig. \@ref(figure2)), which makes it ideal to study the relationships between functional traits and prey specialization. A siphonophore is a colony bearing many feeding polyps (Fig. \@(figure1)), each with a single tentacle, which branches into several tentilla carrying the functional cnidocytes (specialized neural cells carrying nematocysts, the stinging capsules).  Unlike most other cnidarians, siphonophores carry their tentacle nematocysts in extremely complex and organized batteries [@skaer1988formation], built into their tentilla. While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent].

  Many siphonophore species inhabit the deep pelagic ocean, which spans from ~200m to the oceanic seafloor. This habitat has fairly homogeneous physical conditions and stable abundances of zooplanktonic animals [@robison2004deep]. With a somewhat predictable prey availability, ecological theory would predict evolution to drive coexisting siphonophore lineages towards specialization, increasing their feeding efficiencies and reducing interspecific competition [@hardin1960competitive;@hutchinson1961paradox]. If this prediction holds true, we expect the prey capture apparatus morphologies of siphonophores to diversify with the evolution of increased specialization on a variety of prey types in different siphonophore lineages.

  Coexisting siphonophores feeding on the same planktonic community may have substantial niche overlap and compete for prey resources. Traditional ecological coexistence theory [@simpson1944tempo] predicts that competition between species would select for increasing ecological specialization. Moreover, specialization is often thought to be an evolutionary ‘dead end,’ meaning that specialized lineages are unlikely to evolve into generalists or to shift the resource for which they are specialized [@futuyma1988evolution]. However, recent studies have found that interspecific competition can favor the evolution of resource generalism [@stireman2005evolution; @johnson2009competition], and resource switching [@hoberg2008macroevolutionary]. Here we examine three alternative hypotheses on siphonophore trophic specialization: (1) predatory specialists evolved from generalist ancestors; (2) predatory specialists evolved from different ancestral predatory specialists, switching their primary prey type; and (3) predatory generalists evolved from specialist ancestors.

@@

  The few pioneering studies that have addressed the relationships between tentilla and diet suggest that siphonophores are a robust system for the study of predatory specialization via morphological diversification. [@purcell1984functions] and [@purcell1988correlation] showed clear relationships between diet, tentillum, and nematocyst characters in co-occurring epipelagic siphonophores. These correlations, while studied for a small subset of extant epipelagic siphonophore species, might be generalizable to all siphonophores. We hypothesize that these relationships reflect correlated evolution between prey selection and tentillum (and nematocyst) traits. Furthermore, we hypothesize that with an extensive characterization of tentilla morphology, we can generate predictions about the diets of understudied siphonophore species.
In addition, our study design allows us to address other interesting questions about the morphology and evolution of these unique structures. We aim to assess the evolutionary origins of giant tentilla, the phenotypic integration of tentilla, the evolution of the extreme shapes of siphonophore haploneme nematocysts [@thomason1988allometry], and the mechanical implications of tentillum morphologies on cnidoband discharge.

@@

![\label{figure1} Siphonophore anatomy. A - Nanomia sp. siphonophore colony (photo by Catriona Munro). B,C - Illustration of a Nanomia colony, gastrozooid, and tentacle (by Freya Goetz). D - *Nanomia* sp. Tentillum illustration and main parts. E - Transmission micrograph of the tentillum illustrated in D. F - Nematocyst types (illustration reproduced with permission from Mapstone 2014), hypothesized homologies, and locations in the tentillum. Undischarged to the left, discharged to the right.](Figures_final/figure1.jpg)

![\label{figure2} Tentillum diversity plate. The illustrations delineate the pedicle (green), involucrum (blue), cnidoband (orange), elastic strands (pink), terminal structures (yellow). Heteroneme nematocysts (stenoteles in C,E,F,G and mastigophores in H,I) are depicted in red for some species. A - Erenna laciniata, 10x. B - Lychnagalma utricularia, 10x. C - Agalma elegans, 10x. D - Resomia ornicephala, 10x. E - Frillagalma vityazi, 20x. F - Bargmannia amoena, 10x. G - Cordagalma sp., reproduced from Carré 1968. H - Lilyopsis fluoracantha, 20x. I - Abylopsis tetragona, 20x.](Figures_final/figure2.jpg)

@@

*Tentillum morphology* – The morphological work was carried out on siphonophore specimens fixed in 4% formalin from the Yale Peabody Museum Invertebrate Zoology (YPM-IZ) collection (catalog numbers in Appendix 1). These specimens were collected intact across many years of fieldwork expeditions, using of blue-water diving [@haddock2005scientific], remotely operated vehicles (ROVs), and human-operated submersibles. Tentacles were dissected from non-larval gastrozooids, sequentially dehydrated into 100% ethanol, cleared in methyl salicylate, and mounted into slides with Canada Balsam or Permount mounting media. The slides were imaged as tiled z-stacks using differential interference contrast (DIC) on an automated stage at YPM-IZ (with the assistance of Daniel Drew and Eric Lazo-Wasem) and with laser point confocal microscopy using a 488 nm Argon laser that excited autofluorescence in the tissues. Thirty characters (defined in Appendix 2) were measured using Fiji [@collins2007imagej;@schindelin2012fiji]. We did not measure the lengths of contractile structures (terminal filaments, pedicles, gastrozooids, and tentacles), since they are too variable to quantify. We measured at least one specimen for 96 different species (Appendix 3, Fig. \@ref(figure3)). Of these, we selected 38 focal species across clades based on specimen availability and phylogenetic representation. Three to five tentacle specimens from each one of these selected species were measured to capture intraspecific variation.

In order to observe the discharge behavior of different tentilla, we recorded high speed footage (1000-3000 fps) of tentillum and nematocyst discharge by live siphonophore specimens (26 species) using a Phantom Miro 320S camera mounted on a stereoscopic microscope. We mechanically elicited tentillum and nematocyst discharge using a fine metallic pin. We used the Phantom PCC software to analyze the footage. For the 10 species recorded, we measured total cnidoband discharge time (ms), heteroneme filament length (microns), and discharge speeds (mm/s) for cnidoband, heteronemes, haplonemes, and heteroneme shafts when possible (Appendix 4).

*Siphonophore phylogeny* – The phylogenetic analysis included 55 siphonophore species and 6 outgroup cnidarian species (*Clytia hemisphaerica*, *Hydra circumcincta*, *Ectopleura dumortieri*, *Porpita porpita*, *Velella velella*, *Staurocladia wellingtoni*). The gene sequences used in this study are available online (accession numbers in Appendix 5). Some of these sequences we used were accessioned in [@dunn2005molecular], and others we extracted from the transcriptomes in [@munro2018improved]. Two new 16S sequences for *Frillagalma vityazi* (MK958598) and *Thermopalia* sp. (MK958599) sequenced by Lynne Christianson were included and accessioned to NCBI. We aligned these sequences using MAFFT [@katoh2002mafft] (alignments available in Dryad). We inferred a Maximum Likelihood (ML) phylogeny (Appendix 6) from 16S and 18S ribosomal rRNA genes using IQTree [@nguyen2014iq] with 1000 bootstrap replicates (iqtree -s alignment.fa -nt AUTO -bb 1000). We used ModelFinder [@kalyaanamoorthy2017modelfinder] implemented in IQTree v1.5.5. to assess relative model fit. ModelFinder selected GTR+R4 for having the lowest Bayesian Information Criterion score. Additionally, we inferred a Bayesian tree with each gene as an independent partition in RevBayes [@hohna2016revbayes] (Appendix 7 and 9), which was topologically congruent with the unconstrained ML tree. The *alpha* priors were selected to minimize prior load in site variation.

Given the broader sequence sampling of the transcriptome phylogeny, we ran constrained inferences (using both ML and Bayesian timetree approaches, which produced congruent results (Appendix 6 and 7)) fixing the 5 nodes that were incongruent with the topology of the consensus tree in [@munro2018improved]. This tree was then used to inform a Bayesian relaxed molecular clock time-tree in RevBayes, using a birth-death process (sampling probability calculated from the known number of described siphonophore species) to generate an ultrametric tree (Appendix 8). Scripts are available in https://github.com/dunnlab/tentilla_morph/.

*Feeding ecology* – We extracted categorical diet data for different siphonophore species from published sources, including seminal papers [@biggs1977field;@purcell1981dietary;@purcell1984functions;@Mackie:1987uy;@pugh1988two;@bardi2007taxonomic;@andersen1981redescription], and ROV observation data [@choy2017deep;@hissmann2005situ] (Appendix 10). Since the sampling biases of ROV and gut content based methods are very different, the data were merged to the level of feeding guilds. The feeding guilds described here are: small-crustacean specialist (feeding mainly on copepods and ostracods), large crustacean specialist (feeding on large decapods, mysids, or krill), fish specialist (feeding mainly on actinopterygian larvae, juveniles, or adults), gelatinous specialist (feeding mainly on other siphonophores, medusae, ctenophores, salps, and/or doliolids), and generalist (feeding on a combination of the aforementioned taxa, without favoring any one prey group). These were selected to minimize the number of categories without merging the most different types of prey. We removed the gelatinous prey observations for *Praya dubia* eating a ctenophore and a hydromedusa, and for *Nanomia* sp.  eating *Aegina*, since we believe these are rare events that have a much larger probability of being detected by ROV methods than their usual prey, and it is not clear whether the medusae were attempting to prey upon the siphonophores. Personal observations on feeding (from SHDH, CAC, and Philip Pugh)  were also included for *Resomia ornicephala*, *Lychnagalma utricularia*, *Bargmannia amoena*, *Erenna richardi*, *Erenna laciniata*, *Erenna sirena*, and *Apolemia rubriversa*. We extracted copepod prey length data from [@purcell1984functions]. To calculate specific prey selectivities, we extracted quantitative diet and zooplankton composition data from [@purcell1981dietary], matched each diet assessment to each prey field quantification by site, calculated Ivlev’s electivity indices [@jacobs1974quantitative], and averaged those by species (Appendix 11).

*Statistical analyses* – For subsequent comparative analyses, we removed species present in the tree but not represented in the morphology data, and *vice versa*. Although we measured specimens labeled as *Nanomia bijuga* and *Nanomia cara*, we are not confident in some of the species-level identifications, and some specimens were missing diagnostic zooids. Thus, we decided to collapse these into a single taxonomic concept (*Nanomia* sp.). All *Nanomia* sp. observations were matched to the phylogenetic position of *Nanomia bijuga* in the tree. We carried out all phylogenetic comparative statistical analyses in the programming environment R [@team2017r], using the bayesian ultrametric species tree (Fig. \@ref(figure4)), and incorporating intraspecific variation estimated from the specimen data as standard error (Appendix 3). For each character (or character pair) analyzed, we removed species with missing data and reported the number of taxa included. We tested each character for normality using the Shapiro-Wilk test [@shapiro1965analysis], and log-transformed those that were non-normal.

We fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous [@harmon2007geiger]. The models compared were the white noise (WN; non-phylogenetic model that assumes all values come from a single normal distribution with no covariance structure among species), the Brownian Motion (BM) model of neutral divergent evolution [@martins1996phylogenies], the Early Burst (EB) model of decreasing rate of evolutionary change [@harmon2010early], and the Ornstein-Uhlenbeck (OU) model of stabilizing selection around a fitted optimum state [@uhlenbeck1930theory;@butler2004phylogenetic]. We then ranked the models in order of increasing parametric complexity (WN,BM,EB,OU), and compared the corrected Akaike Information Criterion (AICc) support scores [@sugiura1978further] to the lowest (best) score, using a cutoff of 2 units to determine significantly better support. When the best fitting model was not significantly better than a less complex alternative, we selected the least complex model (Appendix 12). We calculated model adequacy scores using the R package arbutus [@pennell2015model] (Appendix 13). We calculated phylogenetic signal in each of the measured characters using Blomberg’s K [@blomberg2003testing] (Appendix 12), and for the morphological dataset as a whole using the R package geomorph [@adams2016geomorph]. We reconstructed ancestral states using Maximum Likelihood (anc.ML [@revell2012phytools]), and stochastic character mapping (make.simmap) for categorical characters. R scripts available in Dryad.

In order to study the evolution of predatory specialization, we reconstructed components of the diet and prey selectivity on the phylogeny using ML (R phytools::anc.ML). To identify evolutionary associations of diet with tentillum and nematocyst characters, we compared the performance of a neutral evolution model to that of a diet-driven directional selection model. First, we collapsed the diet data into the five feeding guilds mentioned above (fish specialist, small crustacean specialist, large crustacean specialist, gelatinous specialist, generalist), based on which prey types they were observed consuming most frequently. We reconstructed the feeding guild ancestral states using the ML function ace (package ape [@paradis2019package]), removing tips with no feeding data. The ML reconstruction was congruent with the consensus stochastic character mapping (Appendix 18). Then, using the package OUwie [@beaulieu2012ouwie], we fitted an OU model with multiple optima and rates of evolution matched to the reconstructed ancestral diet regimes, a single optimum OU model, and a BM null model, inspired by the analyses in [@cressler2015detecting]. Finally, we compared their AICc support values to select the best fitting model (Appendix 14).

@@

To identify the morphological characters with the greatest ability to diagnose and predict diet, we carried out linear discriminant analysis of principal components (DAPC) using the dapc function (R adegenet::dapc) [@jombart2010discriminant]. We generated discriminant functions for feeding guild, soft/hard bodied prey, presence of copepods, fish, and shrimp (large crustaceans) in the diet (Appendix 15). Some taxa have inapplicable states for certain absent characters (such as the length of a nematocyst subtype that is not present in a species), which are problematic for DAPC analyses. We tackled this with two alternative approaches. One approach was to selectively remove characters and taxa to maximize the total data mass. Another approach was to transform the “NA” states to zeroes. The drawbacks of first approach is that useful data are lost, and observations are biased towards taxa with more complex tentacles. The drawback of the second approach is that small state values become aligned with absences. The severity of these drawbacks depends on the specific aspect of diet being predicted, thus we chose our approach in a case-specific basis. When the species with many inapplicable states are removed, we exclude a useful validation set of species (*Apolemia rubriversa*, *Rhizophysa eysenhardtii*, *Rhizophysa filiformis*, and *Physalia physalis*) for which the diet is known. We used the performance of the predictions on these taxa to select the approach used.  For the specific prey type presence in the diet, we then generated and compared generalized linear models (GLMs) with multiple predictors. We chose predictors among the characters with significant loading contributions in each specific DAPC. We then selected the best GLM for each prey type based on the lowest AIC scores.

In order to explore the correlational structure among continuous characters and among their evolutionary histories, we used principal component analysis (PCA) and phylogenetic PCA [@revell2012phytools]. Since the character data contains many gaps due to missing characters, we carried out these analyses on a subset of species and characters that allowed for the most complete dataset. In addition, we obtained the correlations between the phylogenetic independent contrasts [@felsenstein1985phylogenies] using the package rphylip [@revell2014rphylip].

>>>>

---
bibliography: manuscript.bib
csl: systematic-biology.csl
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_document2:
    toc: FALSE
indent: TRUE
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

Alejandro Damian-Serrano^1^^,‡^, Steven H.D. Haddock^2^, Casey W. Dunn^1^

@@

@@

Predators have evolved dedicated body parts to capture and subdue prey. As different predators specialize on distinct prey taxa, their tools for prey capture diverge into a variety of adaptive forms. Studying the evolution of predation is greatly facilitated by a predator clade with structures used exclusively for prey capture that present significant morphological variation. Siphonophores, a clade of colonial cnidarians, satisfy these criteria particularly well, capturing prey with their tentilla (tentacle side branches). Earlier work has shown that extant siphonophore diets correlate with the different morphologies and sizes of their tentilla and nematocysts. We hypothesize that evolutionary specialization on different prey types has driven the phenotypic evolution of these characters. To test this hypothesis, we: (1) measured multiple morphological traits from fixed siphonophore specimens using microscopy and high speed video techniques, (2) built a phylogenetic tree of 45 species, and (3) characterized the evolutionary associations between siphonophore nematocyst characters and prey type data from the literature. Our results show that siphonophore tentillum structure has strong evolutionary associations with prey type and size specialization, and suggest that shifts between prey-type specializations are linked to shifts in tentillum and nematocyst size and shape. In addition, we generated hypotheses about the diets of understudied siphonophore species based on these characters. Thus, the evolutionary history of tentilla shows that siphonophores are an example of ecological niche diversification via morphological innovation and evolution. This study contributes to understanding how morphological evolution has shaped present-day oceanic food-webs.

@@

  Most animal predators have characteristic biological tools that they use to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures tend to be finely attuned to their ability to successfully capture specific prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (*e.g.* avoidance, escape, protective barriers) leads to modifications that can counter those defenses The more specialized the diet of a predator is, the more specialized its tools need to be to meet the specific challenges posed by the prey. Understanding the relationships between predatory specializations and morphological specializations is necessary to contextualize the phenotypic diversity of predators, and to quantify the importance of ecological diversification in generating this diversity.

  Siphonophores (Cnidaria : Hydrozoa) are a clade of organisms bearing modular structures that are exclusively used for prey capture: the tentilla (Fig. \@ref(figure1)). These present a significant morphological variation across species [@mapstone2014global] (Fig. \@ref(figure2)), which makes it ideal to study the relationships between functional traits and prey specialization. A siphonophore is a colony bearing many feeding polyps (Fig. \@ref(figure1)), each with a single tentacle, which branches into several tentilla carrying the functional cnidocytes (specialized neural cells carrying nematocysts, the stinging capsules). Unlike most other cnidarians, siphonophores carry their tentacle nematocysts in extremely complex and organized batteries [@skaer1988formation], built into their tentilla. While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent].

  Many siphonophore species inhabit the deep pelagic ocean, which spans from ~200m to the oceanic seafloor. This habitat has fairly homogeneous physical conditions and stable abundances of zooplanktonic animals [@robison2004deep]. With a relatively predictable prey availability, ecological theory would predict evolution to drive coexisting siphonophore lineages towards specialization, increasing their feeding efficiencies and reducing interspecific competition [@hardin1960competitive;@hutchinson1961paradox]. If this prediction holds true, we expect the prey capture apparatus morphologies of siphonophores to diversify with the evolution of increased specialization on a variety of prey types in different siphonophore lineages.

  Coexisting siphonophores feeding on the same planktonic community may have substantial niche overlap and compete for prey resources. Traditional ecological coexistence theory [@simpson1944tempo] predicts that competition between species would select for increasing ecological specialization. This specialization is often thought to be an evolutionary ‘dead end’, meaning that specialized lineages are unlikely to evolve into generalists or to shift the resource for which they are specialized [@futuyma1988evolution]. However, recent studies have found that interspecific competition can favor the evolution of resource generalism [@stireman2005evolution;@johnson2009competition] and resource switching [@hoberg2008macroevolutionary]. Here we examine three alternative hypotheses on siphonophore trophic specialization: (1) predatory specialists evolved from generalist ancestors; (2) predatory specialists evolved from ancestral predatory specialists which specialized on a different resource, switching their primary prey type; and (3) predatory generalists evolved from specialist ancestors.

@@

  The few pioneering studies that have addressed the relationships between tentilla and diet suggest that siphonophores are a robust system for the study of predatory specialization via morphological diversification. [@purcell1984functions] and [@purcell1988correlation] showed clear relationships between diet, tentillum, and nematocyst characters in co-occurring epipelagic siphonophores. These correlations, while studied for a small subset of extant epipelagic siphonophore species, might be generalizable to all siphonophores. We hypothesize that these relationships reflect correlated evolution between prey selection and tentillum (and nematocyst) traits. Furthermore, we hypothesize that with an extensive characterization of tentilla morphology, we can generate hypotheses about the diets of understudied siphonophore species.
In addition, our study design allows us to address other interesting questions about the morphology and evolution of these unique structures. In particular, we aim to address the evolutionary origins of giant tentilla, the phenotypic integration of tentilla, the evolution of the extreme shapes of siphonophore haploneme nematocysts [@thomason1988allometry], and the mechanical implications of tentillum morphologies on cnidoband discharge.

@@

![\label{figure1} Siphonophore anatomy. A - *Nanomia* sp. siphonophore colony (photo by Catriona Munro). B,C - Illustration of a *Nanomia* colony, gastrozooid, and tentacle (by Freya Goetz). D - *Nanomia* sp. Tentillum illustration and main parts. E - Transmission micrograph of the tentillum illustrated in D. F - Nematocyst types (illustration reproduced with permission from Mapstone 2014), hypothesized homologies, and locations in the tentillum. Undischarged to the left, discharged to the right.](Figures_final/figure1.jpg)

![\label{figure2} Tentillum diversity plate. The illustrations delineate the pedicle (green), involucrum (blue), cnidoband (orange), elastic strands (pink), terminal structures (yellow). Heteroneme nematocysts (stenoteles in C,E,F,G and mastigophores in H,I) are depicted in red for some species. A - *Erenna laciniata*, 10x. B - *Lychnagalma utricularia*, 10x. C - *Agalma elegans*, 10x. D - *Resomia ornicephala*, 10x. E - *Frillagalma vityazi*, 20x. F - *Bargmannia amoena*, 10x. G - *Cordagalma* sp., reproduced from Carré 1968. H - *Lilyopsis fluoracantha*, 20x. I - *Abylopsis tetragona*, 20x.](Figures_final/figure2.jpg)

@@

*Tentillum morphology* – The morphological work was carried out on siphonophore specimens fixed in 4% formalin from the Yale Peabody Museum Invertebrate Zoology (YPM-IZ) collection (accession numbers in Appendix 1). These specimens were collected intact across many years of fieldwork expeditions, using of blue-water diving [@haddock2005scientific], remotely operated vehicles (ROVs), and human-operated submersibles. Tentacles were dissected from non-larval gastrozooids, sequentially dehydrated into 100% ethanol, cleared in methyl salicylate, and mounted into slides with Canada Balsam or Permount mounting media. The slides were imaged as tiled z-stacks using differential interference contrast (DIC) on an automated stage at YPM-IZ (with the assistance of Daniel Drew and Eric Lazo-Wasem) and with laser point confocal microscopy using a 488 nm Argon laser that excited autofluorescence in the tissues. Thirty characters (defined in Appendix 2) were measured using Fiji [@collins2007imagej;@schindelin2012fiji]. We did not measure the lengths of contractile structures (terminal filaments, pedicles, gastrozooids, and tentacles), since they are too variable to quantify. We measured at least one specimen for 96 different species (Appendix 3, Fig. \@ref(figure3)). Of these, we selected 38 focal species across clades based on specimen availability and phylogenetic representation. Three to five tentacle specimens from each one of these selected species were measured to capture intraspecific variation.

In order to observe the discharge behavior of different tentilla, we recorded high speed footage (1000-3000 fps) of tentillum and nematocyst discharge by live siphonophore specimens (26 species) using a Phantom Miro 320S camera mounted on a stereoscopic microscope. We mechanically elicited tentillum and nematocyst discharge using a fine metallic pin. We used the Phantom PCC software to analyze the footage. For the 10 species recorded, we measured total cnidoband discharge time (ms), heteroneme filament length (µm), and discharge speeds (mm/s) for cnidoband, heteronemes, haplonemes, and heteroneme shafts when possible (data in Appendix 4).

*Siphonophore phylogeny* – The phylogenetic analysis included 55 siphonophore species and 6 outgroup cnidarian species (*Clytia hemisphaerica*, *Hydra circumcincta*, *Ectopleura dumortieri*, *Porpita porpita*, *Velella velella*, *Staurocladia wellingtoni*). The gene sequences we used in this study are available online (accession numbers in Appendix 5). Some of the sequences we used were accessioned in [@dunn2005molecular], and others we extracted from the transcriptomes in [@munro2018improved]. Two new 16S sequences for *Frillagalma vityazi* (MK958598) and *Thermopalia* sp. (MK958599) sequenced by Lynne Christianson were included and accessioned to NCBI. We aligned these sequences using MAFFT [@katoh2002mafft] (alignments available in Dryad). We inferred a Maximum Likelihood (ML) phylogeny (Appendix 6) from 16S and 18S ribosomal rRNA genes using IQTree [@nguyen2014iq] with 1000 bootstrap replicates (iqtree -s alignment.fa -nt AUTO -bb 1000). We used ModelFinder [@kalyaanamoorthy2017modelfinder] implemented in IQTree v1.5.5. to assess relative model fit. ModelFinder selected GTR+R4 for having the lowest Bayesian Information Criterion score. Additionally, we inferred a Bayesian tree with each gene as an independent partition in RevBayes [@hohna2016revbayes] (Appendix 7 and 9), which was topologically congruent with the unconstrained ML tree. The *alpha* priors were selected to minimize prior load in site variation.

Given the broader sequence sampling of the transcriptome phylogeny, we ran constrained inferences (using both ML and Bayesian timetree approaches, which produced fully congruent topologies (Appendix 6 and 7)) after fixing the 5 nodes that were incongruent with the topology of the consensus tree in [@munro2018improved]. This topology was then used to inform a Bayesian relaxed molecular clock time-tree in RevBayes, using a birth-death process (sampling probability calculated from the known number of described siphonophore species) to generate ultrametric branch lengths (Appendix 8). Scripts available in Appendix 9.

*Feeding ecology* – We extracted categorical diet data for different siphonophore species from published sources, including seminal papers [@biggs1977field;@purcell1981dietary;@purcell1984functions;@Mackie:1987uy;@pugh1988two;@bardi2007taxonomic;@andersen1981redescription], and ROV observation data [@choy2017deep;@hissmann2005situ] with the assistance of Elizabeth Hetherington and Anela Choy (Appendix 10). We removed the gelatinous prey observations for *Praya dubia* eating a ctenophore and a hydromedusa, and for *Nanomia* sp. eating *Aegina*, since we believe these are rare events that have a much larger probability of being detected by ROV methods than their usual prey, and it is not clear whether the medusae were attempting to prey upon the siphonophores. Personal observations on feeding (from SHDH, CAC, and Philip Pugh) were also included for *Resomia ornicephala*, *Lychnagalma utricularia*, *Bargmannia amoena*, *Erenna richardi*, *Erenna laciniata*, *Erenna sirena*, and *Apolemia rubriversa*. In order to detect coarse-level patterns in the feeding habits, the data were merged into feeding guilds. The feeding guilds described here are: small-crustacean specialist (feeding mainly on copepods and ostracods), large crustacean specialist (feeding on large decapods, mysids, or krill), fish specialist (feeding mainly on actinopterygian larvae, juveniles, or adults), gelatinous specialist (feeding mainly on other siphonophores, medusae, ctenophores, salps, and/or doliolids), and generalist (feeding on a combination of the aforementioned taxa, without favoring any one prey group). These were selected to minimize the number of categories while keeping the most different types of prey separate. We extracted copepod prey length data from [@purcell1984functions]. To calculate specific prey selectivities, we extracted quantitative diet and zooplankton composition data from [@purcell1981dietary], matched each diet assessment to each prey field quantification by site, calculated Ivlev’s electivity indices [@jacobs1974quantitative], and averaged those by species (Appendix 11).

*Statistical analyses* – For subsequent comparative analyses, we removed species present in the tree but not represented in the morphology data, and *vice versa*. Although we measured specimens labeled as *Nanomia bijuga* and *Nanomia cara*, we are not confident in some of the species-level identifications, and some specimens were missing diagnostic zooids. Thus, we decided to collapse these into a single taxonomic concept (*Nanomia* sp.). All *Nanomia* sp. observations were matched to the phylogenetic position of *Nanomia bijuga* in the tree. We carried out all phylogenetic comparative statistical analyses in the programming environment R [@team2017r], using the bayesian ultrametric species tree (Fig. \@ref(figure4)), and incorporating intraspecific variation estimated from the specimen data as standard error (Appendix 3). R scripts available in Dryad. For each character (or character pair) analyzed, we removed species with missing data and reported the number of taxa included. We tested each character for normality using the Shapiro-Wilk test [@shapiro1965analysis], and log-transformed those that were non-normal.

We fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous in the R package *geiger* [@harmon2007geiger]. The models compared were the white noise (WN; non-phylogenetic model that assumes all values come from a single normal distribution with no covariance structure among species), the Brownian Motion (BM) model of neutral divergent evolution [@martins1996phylogenies], the Early Burst (EB) model of decreasing rate of evolutionary change [@harmon2010early], and the Ornstein-Uhlenbeck (OU) model of stabilizing selection around a fitted optimum state [@uhlenbeck1930theory;@butler2004phylogenetic]. We then ranked the models in order of increasing parametric complexity (WN,BM,EB,OU), and compared the corrected Akaike Information Criterion (AICc) support scores [@sugiura1978further] to the lowest (best) score, using a cutoff of 2 units to determine significantly better support. When the best fitting model was not significantly better than a less complex alternative, we selected the least complex model (Appendix 12). We calculated model adequacy scores using the R package *arbutus* [@pennell2015model] (Appendix 13). We calculated phylogenetic signal in each of the measured characters using Blomberg’s K [@blomberg2003testing] (Appendix 12), and for the morphological dataset as a whole using the R package *geomorph* [@adams2016geomorph]. We reconstructed ancestral states using Maximum Likelihood (anc.ML [@revell2012phytools]), and stochastic character mapping (make.simmap) for categorical characters. R scripts available in Dryad.

In order to study the evolution of predatory specialization, we reconstructed components of the diet and prey selectivity on the phylogeny using ML (R phytools::anc.ML). To identify evolutionary associations of diet with tentillum and nematocyst characters, we compared the performance of a neutral evolution model to that of a diet-driven directional selection model. First, we collapsed the diet data into the five feeding guilds mentioned above (fish specialist, small crustacean specialist, large crustacean specialist, gelatinous specialist, generalist), based on which prey types they were observed consuming most frequently. We reconstructed the feeding guild ancestral states using the ML function ace (package ape [@paradis2019package]), removing tips with no feeding data. The ML reconstruction was congruent with the consensus stochastic character mapping (Appendix 18). Then, using the package *OUwie* [@beaulieu2012ouwie], we fitted an OU model with multiple optima and rates of evolution matched to the reconstructed ancestral diet regimes, a single optimum OU model, and a BM null model, inspired by the analyses in [@cressler2015detecting]. Finally, we compared their AICc support values to select the best fitting model (Appendix 14).

@@

To generate hypotheses about the diets of understudied siphonophores for which no feeding observations have yet been reported (but for which we have tentacle morphology data), we carried out linear discriminant analysis of principal components (DAPC) using the dapc function (R adegenet::dapc) [@jombart2010discriminant]. This function allowed us to incorporate more predictors than individuals. We generated discriminant functions for feeding guild, soft/hard bodied prey, presence of copepods, fish, and shrimp (large crustaceans) in the diet (Appendix 15). Some taxa have inapplicable states for certain absent characters (such as the length of a nematocyst subtype that is not present in a species), which are problematic for DAPC analyses. We tackled this by transforming the absent states to zeroes. This approach allows us to incorporate all the data, but creates an attraction bias between small character states (*e.g.* small tentilla) and absent states (*e.g.* no tentilla). Absent characters are likely to be very biologically relevant to prey capture and we believe they should be accounted for. We limited the number of linear discriminant functions retained to the number of groupings in each case. We selected the number of principal components retained using the a-score optimization function (R adegenet::optim.a.score) [@jombart2010discriminant] with 100 iterations, which yielded more stable results than the cross validation function (R adegenet::xval). This optimization aims to find the compromise value with highest discrimination power with the least overfitting. From these DAPCs we obtained the highest contributing morphological characters to the discriminaton (characters in the top quartile of the weighted sum of the linear discriminant loadings controlling for the eigenvalue of each discriminant). For each DAPC we generated hypotheses about the diets of siphonophores outside the training set (R adegenet::predict.dapc), incorporating prediction uncertainty as posterior probabilities (Appendix 15). In order to identify the sign of the relationship between the predictor characters prey type presence in the diet, we then generated generalized logistic regression models (as a type of generalized linear model, or GLM using R stats::glm) with the top contributing characters (from the corresponding DAPC) as predictors. We also carried out these GLMs on the Ivlev’s selectivity indices for each prey type calculated from [@purcell1981dietary] (in Appendix 11).
